Towards Automated Merging of Code Clones in Object-Oriented Programming Languages\\- Work in Progress -

### 2019-05-06 sattose/paper.tex
Sander

\begin{abstract}
Duplication in source code can have a major negative impact on the maintainability of source code, as it creates explicit dependencies between fragments of code. These explicit dependencies have been shown to often cause bugs, as a modification in one fragment might have consequences for the correctness of another. In this study, we look into the opportunities to aid in the process of refactoring these duplication problems for object-oriented programming languages. To do this, we propose a method to be able to detect clones that are suitable for refactoring. This method focuses on taking into account the scope and coupling between expressions in code, to be sure our refactoring improves the design and does not leave side effects after it is applied.

Our intermediate results indicate that more than half of the duplication in code is related to each other through inheritance, making it easier to refactor these clones in a clean way. Approximately a fifth of the duplication can be refactored through method extraction, while other clones require other techniques to be applied. Similar future measurements will provide further insight into the contexts where clones occur, and what this means for the refactoring processes. Finally, we strive to construct a tool that automatically applies refactorings for a large part of the detected duplication problems.
\end{abstract}

@@

Code clones can be found anywhere in a codebase. The location of a clone in the code has an impact on how it has to be refactored. Because of this, we will first look at where clones can be found. Using this information we can determine in which locations clones are found the most, on basis of which the refactoring will be prioritized. A duplicate part in a codebase does not always have to be an exact match with another part to be called a clone. We will look at the definitions of different types of clones and the opportunities to refactor duplicate parts of code albeit not an exact match. We conduct our research on a large corpus of open source projects. We lay the main focus on the Java programming language as refactoring opportunities feature paradigm and programming language dependent aspects~\cite{choi2011extracting}. However, most practices featured in this research will also be applicable to other object-oriented languages, like C\#.

@@

There are many other publications that look into code clone refactoring \cite{alwaqfi2017refactoring, chen2018clone, koni2001scenario}. However, none of the publications describing a refactoring tool have published their source code or application for refactoring support. Furthermore, all of these tools only support a subset of all harmful clones that are found. Also, these publications/tools are limited to suggesting refactoring opportunities for clones found, rather than actually applying refactorings where suitable. Finally, all published approaches have limitations, for instance, false positives when detecting clones \cite{chen2018clone} or being limited to clone pairs \cite{higo2008metric}.

@@

\section{Redefining type 1, 2 and type 3 clones}\label{chap:clonetypes}
In a 2007 survey by Roy et al \cite{roy2007survey} he defined four types of clones:

\textbf{Type 1:} Identical code fragments except for variations in whitespace (may be also variations in layout) and comments.\\
\textbf{Type 2:} Structurally/syntactically identical fragments except for variations in identifiers, literals, types, layout and comments.\\
\textbf{Type 3:} Copied fragments with further modifications. Statements can be changed, added or removed in addition to variations in identifiers, literals, types, layout, and comments.\\
\textbf{Type 4:} Two or more code fragments that perform the same computation but implemented through different syntactic variants.

A higher type of clone means that it's harder to detect. It also makes the clone harder to refactor, as more transformations would be required. Higher clone types also become more disputable whether they actually indicate a harmful antipattern (as not every clone is harmful \cite{jarzabek2010clones, kapser2008cloning}). Type 1 clones are most often, if not always with the right thresholds, harmful. Type 2 and type 3 clones are more disputable whether they are harmful and whether they can be merged in a way that improves the design of the code. In this chapter, we will propose solution for shortcomings in type 1, type 2 and type 3 clones to increase the chance that we can merge the clone while improving the design. Type-4 clones are, because of the serious challenges involved in their detection and refactoring, out of the scope for this study.

\subsection{Type 1 clones} \label{chap:type1clones}
Type 1 clones are identical clone fragments except for variations in whitespace and comments. However, when two clone fragments are textually identical, does not yet indicate that they are actually identical. Even when textually equal, method calls can refer to different methods, type declarations can refer to different types and even variables can be of a different type. This could invalidate a refactoring opportunity for such a code fragment. Because of that, we recommend redefining type-1 clones for refactoring. We propose to:
\begin{itemize}
  \item \textbf{Compare the equality of the fully qualified method signature for method references.} This way we can validate whether two method references, like method calls, are actually equal. In the method signature, not only the fully qualified identifier of the method should be considered, but also the type of all its arguments. This way we can be sure that two potentially cloned method references do not point to overloaded variants (in a case that the data type of arguments is overloaded).
  \item \textbf{Compare the equality of the fully qualified identifier for type references.} This way we can be sure that two referenced types are actually equal, and that they are not just two types with the same name.
  \item \textbf{Compare the equality the fully qualified identifier for variable usages.} Two cloned lines might use a variable with the same name, but different types. This might pose serious challenges on refactoring, as the variables might not concern the same object or primitive. To check this, we need to track the declaration of variables and from this infer the fully qualified identifier of its type.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Considering types:} The definition of type 2 clones states that types should not be considered. We disagree because types can make a significant change to the meaning of a code segment and thus whether this segment should be considered a clone.
  \item \textbf{Having a distinction between different variables:} By the definition of type 2 clones, any identifiers would not be taken into account. We agree that a difference in identifiers may still result in a harmful clone, but we should still consider the distinction between different variables. For instance, if we call a method like this: \textit{``myMethod(var1, var2)''}, or call this method like this: \textit{``myMethod(var1, var1)''}. Even if the variables have the same type, the distinction between the variables is important to ensure the functionality is the same after merging.
  \item \textbf{Defining a threshold for variability in literals:} By the definition of type 2 clones any literals would not be taken into account. We agree, as when merging the clone (for instance by extracting a method), we can simply turn the literal into a method parameter. However, we would argue that thresholds matter here. How many literals may differ for the segment still to be considered a clone with another segment? We need to define a threshold to be sure that, by merging, we are not replacing a code fragment by a worse maintainable design.
  \item \textbf{Consider method call signatures and define a threshold for variability in method calls:} As type-2 clones allow changes in identifiers, also the names of called methods may vary. However, because of this, completely different methods can be called in cloned fragments as a result. This poses serious challenges on refactoring and makes it more disputable whether such a clone is actually harmful. This is because different method identifiers can describe a completely different functionality. Because of that, we recommend to keep considering the call signatures of cloned methods when they are compared. We can allow variability in the rest of method identifiers by passing the function as a parameter. To limit the amount of parameters required we also recommend defining a threshold for variability in method call expressions, so only a limited number of method calls can vary.
\end{itemize}

>>>>

\begin{abstract}
Duplication in source code can have a major negative impact on the maintainability of source code, as it creates explicit dependencies between fragments of code. These explicit dependencies have been shown to often cause bugs, as a modification in one fragment might have consequences for the correctness of another. In this study, we look into the opportunities to automatically refactor these duplication problems for object-oriented programming languages. To do this, we propose a method to detect clones that are suitable for refactoring. This method focuses on taking into account the context and scope of clones, to be sure our refactoring improves the design and does not leave side effects after it is applied.

Our intermediate results indicate that more than half of the duplication in code is related to each other through inheritance, making it easier to refactor these clones in a clean way. Approximately a fifth of the duplication can be refactored through method extraction, while other clones require other refactoring techniques. Similar future measurements will provide further insight into the contexts where clones occur and how this affects the automated refactoring process. Finally, we strive to construct a tool that automatically applies refactorings for a large part of the detected duplication problems.
\end{abstract}

@@

Code clones can be found anywhere in a codebase. The location of a clone in the code has an impact on how it can be refactored. Because of this, we will first look at where clones can be found. Using this information we can determine in which locations clones are found the most, on basis of which the refactoring will be prioritized. A duplicate part in a codebase does not always have to be an exact match with another part to be considered a clone. We will look at the definitions of different types of clones and the opportunities to refactor duplicate parts of code albeit not an exact match. We conduct our research on a large corpus of open source projects. We put the main focus on the Java programming language as refactoring opportunities feature paradigm and programming language dependent aspects~\cite{choi2011extracting}. However, most practices featured in this research will also be applicable to other object-oriented languages, like C\#.

@@

There are many other publications that look into code clone refactoring \cite{alwaqfi2017refactoring, chen2018clone, koni2001scenario}. However, none of the publications describing a refactoring tool have published their source code or application for refactoring support. Furthermore, all of these tools only support a subset of all harmful clones that are found. Also, these publications/tools are limited to suggesting refactoring opportunities for clones found, rather than actually applying refactorings where suitable. Finally, all published approaches have limitations, such as false positives in their clone detection \cite{chen2018clone} or being limited to clone pairs \cite{higo2008metric}.

@@

\section{Redefining type 1, 2 and type 3 clones}\label{chap:clonetypes}
In a 2007 survey by Roy et al \cite{roy2007survey} he defined four types of clones:\\
\textbf{Type 1:} Identical code fragments except for variations in whitespace (may be also variations in layout) and comments.\\
\textbf{Type 2:} Structurally/syntactically identical fragments except for variations in identifiers, literals, types, layout and comments.\\
\textbf{Type 3:} Copied fragments with further modifications. Statements can be changed, added or removed in addition to variations in identifiers, literals, types, layout, and comments.\\
\textbf{Type 4:} Two or more code fragments that perform the same computation but implemented through different syntactic variants.

A higher type of clone means that it's harder to detect. It also makes the clone harder to refactor, as more transformations would be required. Higher clone types also become more disputable whether they actually indicate a harmful antipattern (as not every clone is harmful \cite{jarzabek2010clones, kapser2008cloning}). Type 1 clones are most often, if not always with the right thresholds, harmful. Type 2 and type 3 clones are more disputable whether they are harmful and whether they can be merged in a way that improves the design of the code. In this chapter, we will propose solutions for shortcomings in type 1, type 2 and type 3 clones to increase the chance that we can merge the clone while improving the design. Type-4 clones are, because of the serious challenges involved in their detection and refactoring, out of the scope for this study.

\subsection{Type 1 clones} \label{chap:type1clones}
Type 1 clones are identical clone fragments except for variations in whitespace and comments. However, when two clone fragments are textually identical, it does not yet indicate that they are actually identical. Even when textually equal, method calls can refer to different methods, type declarations can refer to different types and even variables can be of a different type. This could invalidate a refactoring opportunity for such a code fragment. Because of that, we recommend redefining type-1 clones for refactoring. We propose to:
\begin{itemize}
  \item \textbf{Compare the equality of the fully qualified method signature for method references.} If an identifier is fully qualified, it means we specify the full location of its declaration (eg \texttt{com.simonbaars.fruitgame.Apple} for an \texttt{Apple} object). This way we can validate whether two method references, like method calls, are actually equal. In the method signature, not only the fully qualified identifier of the method should be considered, but also the type of all its arguments. This way we can be sure that two potentially cloned method references do not point to overloaded variants (in a case that the data type of arguments is overloaded).
  \item \textbf{Compare the equality of the fully qualified identifier for type references.} This way we can be sure that two referenced types are actually equal, and that they are not just two types with the same name.
  \item \textbf{Compare the equality the fully qualified identifier for variable usages.} Two cloned lines might use a variable with the same name, but different types. This might pose serious challenges on refactoring, as the variables might not concern the same object or primitive. To check this, we need to track the declaration of variables and from this infer the fully qualified identifier of its type.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Considering types:} The definition of type 2 clones states that types should not be considered. We disagree because types can make a significant change to the meaning of a code segment and thus whether this segment should be considered a clone.
  \item \textbf{Having a distinction between different variables:} By the definition of type 2 clones, any identifiers would not be taken into account. We agree that a difference in identifiers may still result in a harmful clone, but we should still consider the distinction between different variables. For instance, if we call a method like this: \texttt{myMethod(var1, var2)}, or call this method like this: \texttt{myMethod(var1, var1)}. Even if the variables have the same type, the distinction between the variables is important to ensure the functionality is the same after merging.
  \item \textbf{Defining a threshold for variability in literals:} By the definition of type 2 clones any literals would not be taken into account. We agree, as when merging the clone (for instance by extracting a method), we can simply turn the literal into a method parameter. However, we would argue that thresholds matter here. How many literals may differ for the segment still to be considered a clone with another segment? We need to define a threshold to be sure that, by merging, we are not replacing a code fragment by a worse maintainable design.
  \item \textbf{Consider method call signatures and define a threshold for variability in method calls:} As type-2 clones allow changes in identifiers, also the names of called methods may vary. However, because of this, completely different methods can be called in cloned fragments as a result. This poses serious challenges on refactoring and makes it more disputable whether such a clone is actually harmful. This is because different method identifiers can describe a completely different functionality. Because of that, we recommend to keep considering the call signatures of cloned methods when they are compared. We can allow variability in the rest of method identifiers by passing the function as a parameter. To limit the amount of parameters required we also recommend defining a threshold for variability in method call expressions, so only a limited number of method calls can vary.
\end{itemize}
